Working Title: Roles and Expectations in Software Organizations

### 2011-09-26 soulgem.tex
Second full draft (including abstract) complete.

\begin{document}
\title{The Soulgem Paper}

@@

Let me not to the marriage of true minds / Admit impediments, love is not love / Which alters when it alteration finds, / Or bends with the remover to remove.

O no, it is an ever fixed mark / That looks on tempests and is never shaken; / It is the star to every wand'ring bark, / Whose worth's unknown although his height be taken.

Love's not time's fool, though rosy lips and cheeks / Within his bending sickle's compass come, / Love alters not with his brief hours and weeks, / But bears it out even to the edge of doom:

If this be error and upon me proved, / I never writ, nor no man ever loved.
\end{abstract}

\begin{IEEEkeywords}
pink robots;
\end{IEEEkeywords}

@@

The following points summarize the main elements of the previous discussion.

@@

We expected to find that people enacting similar roles would use similar communication media, and that they would communicate with people enacting interfacing roles through other media. Although we found support for this claim, as we explain below, a biasing factor was the prevalence of face-to-face interactions in both of our field sites. Face-to-face was widely used in communication between people with similar and with different roles. Stand-up meetings, walks across the hall, and group meetings were commonplace.

@@

% TODO: Need to add *bolded* sentences to point out all of the findings in each section.

@@

We explored the sociological and organizational literature to generate a theoretical framework of roles and expectations in software organizations. Using that framework we studied several research questions regarding work in software groups.

We found that roles, defined as sets of expectations, were inconsistent within and between the organizations we studied. Roles are socially constructed; part of the construction occurs at the cultural or institutional level, and part occurs at the local, team level, and is caused by historical, structural, and personal factors. Therefore, although we can make some reasonable assumptions about the work of a software professional based on his role label or official position, these assumptions will only take us so far: professionals vary considerably in their responsibilities, and treating them as archetypical professionals (``developers,'' ``testers,'' and so on) dismisses a large part of what they bring or fail to bring to their job.

@@

We also explored role evolution and found that it is driven by organizational growth (which leads to specialization) and by mismatches between expectations and behaviour (which lead to a readjustment of expectations). Finally, we found that people enacting different roles tend to use different media to communicate, and that face-to-face interaction and issue-tracking notifications were the most widely used media in our sites in terms of inter-role communications, with the qualification that issue-tracking systems were used differently by people enacting different roles.

@@

\emph{Research:} First, the fact that role definitions are not institutionally uniform and respond to local (historical, structural, personal) factors should help clarify common misunderstandings and assumptions made by researchers in our studies. An archetypical developer, team lead, or tester, is rarely found in practice. Instead, thinking in terms of tasks and the people who are expected to perform them will help us reattach our studies and findings to software development in the real world.

Second, thinking in terms of sets of expectations, rather than positions, clears the path for an analysis of good patterns of expectation clusters (that is, the expectations that make good sense, organizationally, to be placed on the same individual), which should lead to better role definitions, and for the study of bottlenecks and deficiencies in organizational structure. For instance, the discussion of the proper ratio of developers to testers, which is moot in the case of teams where there is no QA and developers do all the testing, can be now explored in terms of the proper ratio of people expected to write production code to people expected to write integration tests or to determine the level of quality of the product.

Third, the observation that media use varies depending on the kinds of expectations placed on the user complements and helps explain the earlier finding that electronic repositories are incomplete and often misleading \cite{Aranda2009}. If software repository miners focus on a single electronic repository, they are likely to miss systematically important areas of organizational coordination: they will be blind to some people and expectations. Depending on the research questions they address, this may be a serious concern. Mining issue tracking systems, judging by our two cases, should yield the greatest payoff, although even then, mining algorithms would miss the people that mostly listen or pay attention to the aggregate, and focus only on those who emit messages in the system.

@@

Additionally, we found that many practitioners do not have an awareness of the full scope of their peers' activities, and that such an awareness leads them to misunderstand the benefits that they bring to the organization. It may be the case for them, as it was in the two groups we studied, that people nominally tasked with minor matters actually help the team in deeper, more complex ways, and that the their formal position does not match the expectations that various groups of people have of them. In these cases a reductionism to position labels would be harmful.

@@

The two organizations we studied were small (less than fifty members at the time of our observations), geographically co-located for the most part, and North American. The extent to which our findings are generalizable might therefore be questioned. However, most of our findings are supported by a theoretical framework that is not built upon these particular characteristics, but is instead general and interdisciplinary. We expect that some of the particular expectation clusters will vary in other organizations, depending on their size, physical co-location, and culture, but we also expect the underlying rationale and observations to hold across them.

@@

The framework and our findings generate several further questions, which we plan to address in future work. First, there is the issue of the right level of granularity to articulate expectations in software development. Second, an analysis of commonly-found and commonly-useful patterns in expectation clustering. Third, an analysis of  appropriate ratios of expectation assignment in software groups. And fourth, an exploration of the differences in specialization and expectation clustering in organizations that differ significantly from those we observed (that is, in larger organizations, geographically distributed, and imbued in different cultures). We hope others in the research community join us in addressing these questions.

>>>>

\begin{document}
\title{Working Title: Roles and Expectations in Software Organizations}

@@

Software professionals divide their work by roles, but the nature of these roles has not been examined nor questioned in our field. In this paper we formulate a theoretical framework to address this issue. The framework is based on sociological and organizational studies, and defines roles as sets of expectations placed on the members of the software organization.

The paper reports on an empirical case study of two software organizations, which was informed and inspired by the theoretical framework. In our study we found that roles were inconsistent within and between the organizations we studied, and that role definitions are shaped by institutional, historical, structural, and personal factors. Furthermore, role definitions evolve through organizational growth and through mismatches between expectations and performance, and people enacting different roles use different media to communicate.

Our contributions should help researchers to frame better studies of division of labour in software groups, and to avoid the pitfalls that come from unwarranted abstractions of software professional roles. They also provide practitioners with a conceptual tool to reason about social structure and organizational design.

\end{abstract}

\begin{IEEEkeywords}
software organizations; social structure; roles
\end{IEEEkeywords}

@@

@@

We expected to find that people enacting similar roles use similar communication media, and that they communicate with people enacting interfacing roles through other media. Although we found support for this claim, as we explain below, a biasing factor was the prevalence of face-to-face interactions in both of our field sites. People with similar or different roles communicate face-to-face regularly. Stand-up meetings, walks across the hall, and group meetings were commonplace.

@@

@@

We explored the sociological and organizational literature to generate a theoretical framework of roles and expectations. We used that framework to study several research questions regarding work in software organizations.

We found that roles, defined as sets of expectations, were inconsistent within and between the organizations we studied. Roles are socially constructed; part of the construction occurs at the cultural or institutional level, and part occurs at the local, team level, and is caused by historical, structural, and personal factors. Although we can make some reasonable assumptions about the work of a software professional based on his label or official position, our assumptions will only take us so far: their responsibilities vary considerably, and treating them as archetypical professionals (``developers,'' ``testers,'' and so on) dismisses a large part of what they bring or fail to bring to their job.

@@

We also explored role evolution and found that it is driven by organizational growth (which leads to specialization) and by mismatches between expectations and behaviour (which lead to an adjustment of expectations). Finally, we found that people enacting different roles tend to use different media to communicate, and that face-to-face interaction and issue-tracking notifications were the most widely used media in our sites in terms of inter-role communications, with the qualification that issue-tracking systems were used differently by people enacting different roles.

@@

\emph{Research:} First, the fact that role definitions are not institutionally uniform and respond to local (historical, structural, personal) factors should help clarify common misunderstandings and assumptions made in research. A stereotypical developer, team lead, or tester, is rarely found in practice. Instead, thinking in terms of tasks and the people who are expected to perform them will help us reattach our studies and findings to software development in the real world.

Second, thinking in terms of sets of expectations, rather than positions, clears the path for an analysis of good patterns of expectation clusters (that is, the expectations that make good sense, organizationally, to be placed on the same individual), which should lead to better role definitions, and for the study of bottlenecks and deficiencies in organizational structure. For instance, the discussion of the proper ratio of developers to testers, which is moot for teams where there is no QA and developers do all the testing, can be now explored in terms of the proper ratio of people expected to write production code to people expected to write integration tests or to determine the level of quality of the product.

Third, the observation that media use varies depending on the kinds of expectations placed on the user complements and helps explain the earlier finding that electronic repositories are incomplete and often misleading \cite{Aranda2009}. If researchers focus on a single electronic repository, they are likely to miss systematically important areas of organizational coordination: they will be blind to some people and expectations. Depending on the research questions they address, this may be a serious concern. Mining issue tracking systems, judging by our two cases, should yield the greatest payoff, although even then, mining algorithms would miss the people that mostly listen or pay attention to the aggregate, and focus only on those who emit messages in the system.

@@

Additionally, we found that many practitioners do not have an awareness of the full scope of their peers' activities, and that this lack of awareness leads them to misunderstand the benefits that they bring to the organization. It may be the case for them, as it was in the two groups we studied, that people nominally tasked with minor matters actually help the team in deeper, more complex ways, and that their formal position does not match the expectations placed on them. In these cases a reductionism to position labels is harmful.

@@

The two organizations we studied were small (less than fifty members), geographically co-located for the most part, and North American. Therefore, the generalizability of our findings might be questioned. However, most of our findings are supported by a theoretical framework that is not built upon these particular characteristics, but is instead general and interdisciplinary. We expect that some of the particular expectation clusters will vary in other groups, depending on their contextual factors, but we also expect the underlying rationale and observations to hold across them.

@@

The framework and our findings generate several further questions, which we plan to address in future work. First, there is the issue of the right level of granularity to articulate expectations in software development. Second, an analysis of commonly-found and commonly-useful patterns in expectation clustering. Third, an analysis of  appropriate ratios of expectation assignment in software groups. And fourth, an exploration of the differences in specialization and expectation clustering in organizations that differ significantly from those we observed. We hope others in the research community join us in addressing these questions.
